### 2019-07-15 Thesis/intro.tex
Daily commit.

We can then use concepts from category theory to inform our denotational semantics. 
The most important of these is that if the category we choose is Cartesian closed, then we can automatically interpret terms of the simply typed lambda calculus as morphisms.
This is particularly useful, since many of the languages that we are most interested in studying from a mathematical point of view -- in particular, PCF and Idealized Algol -- are based upon the simply typed lambda calculus.
Thus, if we make sure that our base category is Cartesian closed, then we get a denotational semantics for the lambda calculus-part of the language for free, allowing us to concentrate on the other parts of the language.

Category-theoretic topics have taken on a new importance within functional programminging.
In 1991, Eugenio Moggi observed that \emph{monads} on categories provide a way to study computational effects. 
For example, we have already alluded to the fact that a stateful function $A \to B$ may be identified with an ordinary stateless function
\[
	A \times W \to B \times W\,,
	\]
where $W$ is some fixed set (the \emph{state}). 
Such a function may alternatively be written as a function
\[
	A \to (W \to (B \times W))\,;
	\]
thus, a morphism $A \to B$ in the Kleisli category for the \emph{state monad}
\[
	B \mapsto W \to (B \times W)\,.
	\]
Similarly, a nondeterministic function $A\to B$ may be thought of as a function from elements of $A$ to subsets of $B$, and thus as a morphism in the Kleisli category for the \emph{powerset monad} $\powerset-$.

The link between monads and computational effects has not been confined to the theoretical domain; indeed, monads quickly became the main way of expressing effectful programming in languages such as Haskell \cite{Haskell}.

In 2001, Plotkin and Power \cite{PlotkinPower} made this idea more precise by demonstrating that if a computational effect can be given by algebraic operations and equations, then it determines a monad.

More specifically, a \emph{signature} $\Sigma$ is made up of a set of \emph{operations} $\sigma$ with (possibly infinite) \emph{arities} $\text{ar}(\sigma)$ and \emph{equations} that encode the rules that hese operations must follow.
An \emph{interpretation} of a signature $\Sigma$ in a category $\C$ with products is given by an object $A$ of $\C$, together with morphisms
\[
	A^\text{ar}(\sigma) \to A
	\]
for each operation $\sigma$ in $\Sigma$, such that the equations all hold when interpreted as equalities of morphisms in $\C$.

A \emph{homomorphism} between $\Sigma$-algebras $A$ and $B$ is a morphism $A \to B$ that respects the $\Sigma$-algebra structure on $A$ and $B$ in a natural way. 
Thus, we get a category of $\Sigma$-algebras.
Moreover, the natural forgetful functor from this category into $\C$ admits a left adjoint, inducing a monad on $\C$.
This monad sends a set $A$ to the underlying set of the \emph{free $\Sigma$-algebra} on $A$, given by taking the set of all terms that can be formed from the operations in $\Sigma$, with the elements of $A$ as variables, and quotienting by the equivalence generated by the equations in $\Sigma$.

For example (see \cite{Powerset}, for instance), the powerset monad for countable sets is induced from the signature $\Sigma_\powerset$ consisting of a single infinitary operator $\sum$ (i.e., nondeterministic choice between the infinite collection of values), together with equations for infinitary commutativity (reordering the $a_i$ should not change the observed behaviour), associativity ($\sum_i \sum_j a_{ij} = \sum_{i,j} a_{ij}$) and idempotence (repeat entries do not change the behaviour).

If $A$ is a set, then we may identify an element $\phi$ of the free $\Sigma_\powerset$-algebra on $A$ with the subset of $A$ consisting of all elements $a\in A$ that appear in the term-description of $\phi$.
For example, if $A=\bN$, then we can identify
\[
	\sum_i 2i
	\]
with the set of even natural numbers, and
\[
	\sum_i \sum_j \pi(i)^j
	\]
with the set of all prime powers.
The axioms for commutativity, associativity and idempotence mean that this is a well-defined operation, giving us an equivalence between the free $\Sigma_\powerset$-algebra monad and the powerset monad $\powerset$.

\newcommand{\lku}{\operatorname{lookup}}
\newcommand{\upd}{\operatorname{update}}

The state monad with state given by a set $W$ is induced from the signature given by a $W$-ary operation $\lku$, corresponding to choosing between $W$-many options based on the value of the state; and $W$-many nullary operations $\upd_w$, each corresponding to updating the state to a particular value $w$.
There are four axioms relating these two operations, which we adapt from \cite{PlotkinPower}.
\begin{description}
	\item[Update with current value] 
          \[
            \lku(\upd_w(x)\suchthat w\in W) = x\,.
            \]
          Looking up the current value and rewriting it is a no-op.
	\item[Idempotence of lookup]
          \[
            \lku(\lku(x_{vw}\suchthat v\in W)\suchthat w\in W)=\lku(x_{ww}\suchthat w\in W)\,.
            \]
          If we look up the state twice in succession, then we will get the same value both times.
	\item[Double update] 
          \[
            \upd_v(\upd_w(x))=\upd_w(x)\,.
            \]
          If we update the value twice in succession, then the first update has no effect.
	\item[Lookup] 
          \[
            \upd_w(\lku(x_v\suchthat v\in W)) = \upd_w(x_w)\,.
            \]
          If we update the value to $w$ and then look up the state, we get $w$ back.
\end{description}

Plotkin and Power prove in \cite{PlotkinPower} that the free algebra monad for this signature is isomorphic to the usual state monad $(W \times -)^W$.
We therefore get a strong connection between some very concrete syntactic concepts about what state should look like and the abstract category-theoretic concept of a monad.

>>>>

Typically, each type $T$ of the language will correspond to some object $\deno T$ of the category, while a term of type $T$ will correspond to a morphism $1 \to \deno T$, where $1$ is some fixed object in the category (usually a terminal object).

Particularly important \cite{Lambek} are the Cartesian closed categories, which have a number of properties making them suitable for denotational semantics:
\begin{description}
  \item[Product and function spaces] Given types $S$ and $T$, we can define the denotations of the product type $S \times T$ and the function type $S \to T$ to be given by $\deno S \times \deno T$ and ${\deno T}^{\deno S}$.
  \item[Compositionality] Given types $S$ and $T$, and corresponding objects $\deno S$ and $\deno T$ of the category, we can define the denotation of the function type $S \to T$ to be given by the exponentiation ${\deno T}^{\deno S}$ as above.
    Then we automatically have a recipe for substituting a term of type $S$ into a function of type $S\to T$ via the canonical morphism
    \[
      {\deno T}^{\deno S} \times \deno S \to \deno T\,.
      \]
  \item[Abstraction] Given a morphism $\sigma \from A \times B \to C$, we may form a morphism $\Lambda(\sigma) \from A \to C^B$.
    This gives us the semantics for $\lambda$-abstraction, whereby we pass from a term-in-context
    \[
      \Gamma,x\from S \ts M \from T
      \]
    to the term-in-context
    \[
      \Gamma \ts \lambda x.M\from S \to T\,.
      \]
\end{description}
These rules allow us to build up a model of the simply-typed $\lambda$-calculus within any Cartesian closed category, which means we get a large part of the denotation (and the subsequent proof of Computational Adequacy) for free.

This alone would be a good justification for using category theory in denotational semantics, but the benefits go further.  
The development of programming languages such as Haskell has been strongly influenced by category-theoretic concepts.  
For example, Moggi's 19xx observation \cite{Moggi} that monads on categories provide us with a way of modelling computational effects influenced work that led directly to the introduction of support for monads in Haskell \cite{Haskell}, where they have become the primary tool for abstracting out effectful computation.

Monads will be particularly important in this thesis, so it is worth dwelling on them a little further.  
A \emph{monad} on a category $\C$ is given by a functor $M\from \C \to \C$, together with natural transformations
\begin{mathpar}
  e\from\id_\C \Rightarrow M \and m\from M\circ M\Rightarrow M
\end{mathpar}
that endow $M$ with an algebraic structure.
One example is the non-empty powerset functor on the category of sets, together with the natural transformations given by
\begin{mathpar}
  \begin{IEEEeqnarraybox*}{rCcCl}
    e & \from & A & \to & \powerset(A) \\
    && a & \mapsto & \{a\}
  \end{IEEEeqnarraybox*}
  \and
  \begin{IEEEeqnarraybox*}{rCcCl}
    m & \from & \powerset(\powerset(A)) & \to & \powerset(A) \\
    && \mathcal A & \mapsto & \bigcup_{A \in \mathcal A} A\,.
  \end{IEEEeqnarraybox*}
\end{mathpar}
This powerset monad indicates some kind of nondeterministic choice between elements of $A$, particularly if we modify the construction to the \emph{non-empty powerset} functor $\powerset_+$.

Another example in the category of sets is the functor $A \mapsto A + \{\bot\}$, that appends an additional element on to a set.  
We have natural functions $A \to A + \{\bot\}$ and $A + \{\bot\} + \{\bot\} \to A + \{\bot\}$ that make this into a monad as well.
In the study of programming languages, this is often called the \emph{maybe monad}, because $A + \{\bot\}$ indicates an element of $A$ that may or may not be present (with the distinguished value $\bot$ indicating no value).

Given a monad $M$ on a category $\C$, we can form a new category $\Kl_M\C$ -- the \emph{Kleisli category} of $M$ -- whose objects are the objects of $\C$ and where a morphism from $A$ to $B$ is given by a morphism $A \to MB$ in $\C$.  
The monadic coherence gives us the correct notion of composition: given Kleisli morphisms $\sigma \from A \to MB$ and $\tau\from B \to MC$, we may compose them to give a morphism $A \to MC$ via the following formula.
\[
  A \xrightarrow{\sigma} MB \xrightarrow{M\tau} M M C \xrightarrow{m} MC
  \]
The Kleisli category of the powerset monad is the category of sets and relations, while the Kleisli category of the maybe monad is the category of sets and partial functions.

There are numerous other monads that can be used to model computational effects, such as the state monad and the exception monad.  
Work by Plotkin and Power \cite{PlotkinPower} makes this more precise, by studying monads that can be built up via algebraic operations and equations.  
For example, we might want to model nondeterministic choice on a set $A$ via an operation $\sum$ that takes in infinitely many elements of $A$ -- so $\sum a_i$ gives us a choice between the $a_i$.  
We then impose some axioms on this operation.
\begin{description}
  \item[Idempotence] If $a_i=a$ for all $i$, then $\sum a_i=a$;
  \item[Commutativity] $\sum a_i = \sum_{a_{\pi(i)}}$ for any permutation $\pi$; and
  \item[Associativity] $\sum_i (\sum_j a_{ij}) = \sum_{i,j} a_{ij}$.
\end{description}
This is an algebraic theory akin to the theory of groups, and its category of free algebras is isomorphic to the Kleisli category of the powerset monad.

\section{Game Semantics}

Game Semantics gives us a particularly fruitful categorical semantics for programming languages.  
The underlying idea is that a computer program behaves like a strategy for a two-player game, in that it needs to respond to arbitrary inputs (opponent moves) with its own behaviours (proponent moves).  
Thus, we represent a programming language type by an idealized two-player game and represent a term of that type by a strategy for that game.

The precise definition of a strategy that we use depends on the language that we are trying to model -- a language with less expressive power can realize fewer strategies.  
For example, the denotations of terms in a stateless langauge such as PCF are \emph{history-free} or \emph{innocent}, in which the proponent's moves can only depend on a particular subsequence of the current sequence of moves -- the $P$-view -- rather than on the whole sequence.
So for a lot of computational effects, particularly ones that have something to do with state, adding that effect corresponds to a \emph{relaxing} of conditions on strategies.

We model other types of effects by extending the definition of a strategy.  
For example, if we want to provide a semantics for a language with nondeterminism, then we modify the definition of a strategy so that the propoonent can have multiple replies to each opponent move, as in the work of Harmer and McCusker \cite{mcCHFiniteND}.
If we want to model a probabilistic language, then we decorate these different moves with probabilities, as in the work of Danos and Harmer \cite{DanosHarmer}.

When choosing a definition of a strategy, the aim is to prove a definability result, so that we can prove Full Abstraction.  
The original proofs of definability of compact innocent strategies in PCF from \cite{ajmPcf} and \cite{hoPcf} were intricate and technical.  
Subsequent work on languages that extend PCF tends to try to prove definability via a \emph{factorization result}, in which we show that every strategy in an extended category of games may be written as the composition of a strategy in an original category of games with some fixed strategy in the new category.  
Then, if we have a definability result for the original semantics, we can extend it to a definability result in the new category.

For example, the language Idealized Algol is an extended version of PCF that adds some stateful primitives.
For example, Abramsky and McCusker's proof of compact definability for Idealized Algol in \cite{SamsonGuyIAActive} first proves that each compact strategy in their model may be written as the composite of a compact innocent strategy with the (non-innocent) denotation of one of the new stateful constants.  
Thus, they can deduce compact definability for Idealized Algol from Hyland and Ong's result that every compact innocent strategy is the denotation of a term of PCF.

Harmer and McCusker develop in \cite{mcCHFiniteND} a model of game semantics in which strategies can be nondeterministic.  
They show that every such strategy can be written as the composite of a deterministic strategy with some particular fixed nondeterministic strategy.  
Then, to prove compact definability for nondeterministic Idealized Algol, it suffices for them to exhibit a nondeterministic term whose denotation is that fixed strategy.
